A generalization of the Bradley-Terry model for draws in chess

### 2019-05-16 inst/collusion_chess_draws.Rnw
work

\newcommand{\oo}{\vphantom{W^{W^W}}}

@@

Here I analyse the original subject of the complaint, the Cura\c{c}ao
1962 Candidates, Table~\ref{candidates}.  Fischer alleged that three
of the Soviet players (Keres, Petrosian, and Geller) had agreed to
draw amongst themselves.

%the question of Soviet collusion during the period when the USSR
%dominated the struggle for the world championship has been one of the
%long-standing debates in the chess world.  The best-known (but not
%unique) criticism was aired by Bobby Fischer some weeks after the 1962
%Candidates Tournament in Curacao

@@

@@

\begin{table}
\centering
\begin{tabular}{l|cccccc}
&\rotatebox{-90}{Fischer} 
&\rotatebox{-90}{Geller} 
&\rotatebox{-90}{Petrosian\ }
&\rotatebox{-90}{Korchnoi} 
&\rotatebox{-90}{Filip}
& $\cdots$
\\ \hline %       Fischer   Geller    Petrosian Korchnoi    Filip
Fischer    (USA)  &   -     & \halfb  & \halfw  & \oneww    &\halfb    &\\
Geller    (USSR)  & \halfw  &    -    & \halfb  & \halfw    & \halfw   &\\
Petrosian (USSR)  & \halfb  & \halfw  &    -    & \halfb    & \halfw   &\\
Korchnoi  (USSR)  & \zerob  & \halfb  & \halfw  &   -       & \oneww   &\\
Filip      (TCH)  & \halfw  & \halfb  & \halfb  & \zerob    &  -       &\\
$\vdots$&&&&&\\
\end{tabular}\label{interzonal}
\caption{Subset of the Fifth Interzonal round-robin, Stockholm 1962.
  Results $\in\{0,\halfw,1\}$ and box colour refer to row
  player.  Thus, for example, Fischer played black against Geller and drew}
\end{table}

>>>>

\newcommand{\oo}{\vphantom{W^{W^{W^W}}}}

@@

@@

Here I analyse the original subject of the complaint, the Cura\c{c}ao
1962 Candidates, Table~\ref{candidates}.  Fischer alleged that three
of the Soviet players (Keres, Petrosian, and Geller) had agreed to
draw amongst themselves.  Observe that all twelve of the games between
the three accused players ended in a draw, a significant result given
51 winning games and 54 draws (Fisher exact test, one-sided,
$p=0.0079$).  The likelihood function for this dataset would have

\begin{equation}
\parbox{4in}{$
P^5\,
{Ke}^5\,
G^4
\mathit{Fis}\,  % Fischer, cf Filip
\mathit{Ko}^5
B
T^2  % Tal
\mathit{Fil}\,
D^{42}C^{12}\oo\\ \rule{10mm}{0mm}
\left(P+W\right)^3
\left(\mathit{Ke}+W\right)^4
\left(G+W\right)^4\
\left(\mathit{Fis}+W\right)^7\oo\\ \rule{20mm}{0mm}
\left(\mathit{Ko}+W\right)^2
\left(B+W\right)^5
\left(T+W\right)
\left(\mathit{Fil}+W\right)\oo
$}
\end{equation}

on the numerator (where $C$ is the strength of a collusive draw), and
a total of 36 terms such as $\left(P + \mathit{Fis}+W+D\right)^{4}$
and $\left(P + \mathit{Ke}+W+C\right)^{4}$ on the denominator.

@@

\begin{table}
\centering
\begin{tabular}{l|cccccc}
&\rotatebox{-90}{Fischer} 
&\rotatebox{-90}{Geller} 
&\rotatebox{-90}{Petrosian\ }
&\rotatebox{-90}{Korchnoi} 
&\rotatebox{-90}{Filip}
& $\cdots$
\\ \hline %       Fischer   Geller    Petrosian Korchnoi    Filip
Fischer    (USA)  &   -     & \halfb  & \halfw  & \oneww    &\halfb    &\\
Geller    (USSR)  & \halfw  &    -    & \halfb  & \halfw    & \halfw   &\\
Petrosian (USSR)  & \halfb  & \halfw  &    -    & \halfb    & \halfw   &\\
Korchnoi  (USSR)  & \zerob  & \halfb  & \halfw  &   -       & \oneww   &\\
Filip      (TCH)  & \halfw  & \halfb  & \halfb  & \zerob    &  -       &\\
$\vdots$&&&&&\\
\end{tabular}
\caption{Subset of the Fifth Interzonal round-robin, Stockholm 1962.\label{interzonal}
  Results $\in\{0,\halfw,1\}$ and box colour refer to row
  player.  Thus, for example, Fischer played black against Geller and drew}
\end{table}
